\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,booktabs,graphicx,microtype,url,xcolor}
\usepackage[hidelinks]{hyperref}
\title{Vulkax Atlas: Route-Predictive Joint Resource Allocation\\
for a Vulkan Globe and Navigation Runtime}
\author{Swayam Singal}
\date{July 2026}

\begin{document}
\maketitle

\begin{abstract}
Planet-scale navigation renderers must place useful content in memory before the camera reaches it,
while simultaneously controlling frame time, upload bandwidth, geometric detail, and lighting
error. Visibility- or distance-only streaming reacts to the current view but ignores a route's
strong prediction of future demand. This paper presents Vulkax Atlas, a C++20 Vulkan globe and
navigation research runtime built on BEACON and GeoBEACON. Atlas combines double-precision WGS84
and Earth-centered Earth-fixed coordinates, six-face cube-quadtree selection, cancellable
file/HTTP/SQLite streaming, offline regional packs, normalized search/routing/transit/traffic
providers, and a route-predictive budget controller. The controller allocates tile residency,
upload bandwidth, geometric utility, and bounded diffuse-light work using route probability,
time-to-arrival, semantics, visibility, and measured frame pressure. A deterministic constrained
scheduler experiment produces raw frame rows and seven ablations. In this synthetic stress fixture,
distance-only and geometry-only policies each recorded 644 route-corridor deadline misses,
velocity-only recorded 204, route-only recorded 8, and route-plus-semantics and full Atlas each
recorded 4. These values are explicitly classified as analytical scheduler simulation rather than
Vulkan performance measurements. The artifact preserves the checked GeoBEACON renderer, exact
diffuse reference captures, Vulkan query-pool timing, five reproducible region manifests, and
per-source attribution.
\end{abstract}

\section{Introduction}
A globe renderer has a future-information problem. The current camera identifies visible content,
but a turn-by-turn route identifies content likely to become visible seconds later. A naive
streamer either requests too late, creating low-detail transitions and missing labels, or requests
too broadly, wasting bandwidth and memory. The same route also changes the value of rendered
content: an upcoming intersection, maneuver label, station, or landmark matters more than an
equally sized object outside the route corridor.

Vulkax Atlas tests the following hypothesis:
\begin{quote}
\emph{Joint route- and semantic-aware allocation of geometry detail, streaming bandwidth, memory,
and lighting accuracy reduces route-corridor deadline misses or wasted prefetch bytes compared with
distance-only and velocity-only control under an equal budget.}
\end{quote}
The contribution is the unified allocation policy and its ablations. Globe rendering, cube maps,
screen-space error (SSE), clustered lighting, route engines, and HTTP caching are supporting
systems rather than novelty claims.

Atlas extends GeoBEACON from one checked neighborhood to a planet-ready coordinate and content
contract. Existing BEACON technique identifiers and GeoBEACON datasets remain regression baselines.
The renderer distinguishes three evidence classes: Vulkan GPU timestamps, Vulkan CPU fallback, and
analytical scheduler simulation. Conservative omitted-light bounds also remain separate from
rendered MSE, PSNR, SSIM, and maximum pixel error.

\section{Prior Work and Positioning}
Hierarchical terrain and globe systems routinely use quadtree refinement and SSE. OGC 3D Tiles 1.1
standardizes hierarchical spatial content, implicit tiling, glTF payloads, feature identifiers, and
structural metadata~\cite{tiles}. OpenStreetMap (OSM) provides global vector geometry and replication
diffs under the ODbL~\cite{osm,replication}. Pelias, Valhalla, and OpenTripPlanner provide open
geocoding, road routing, and multimodal transit foundations~\cite{pelias,valhalla,otp}.

Game engines also stream world sectors using visibility, distance, predicted movement, authored
importance, and mission state. Atlas does not claim novelty for predictive streaming in isolation.
Its research claim is narrower: one transparent controller combines route probability,
time-to-arrival, maneuver and semantic importance, geometry and upload cost, and BEACON's measured
lighting/error controls; each signal is independently disabled in a reproducible ablation.

Clustered and tiled shading established spatial light lists. Lightcuts-style systems established
bounded many-light approximation, while grid-based reservoirs such as ReGIR target much larger
stochastic many-light workloads~\cite{regir}. Atlas retains BEACON's deterministic finite-radius
point-light scope. Formal error accounting covers opaque, unshadowed diffuse direct light; visual
specular and shadows are outside the guarantee.

\section{Planetary Coordinate Model}
Persistent positions use WGS84 latitude $\phi$, longitude $\lambda$, and ellipsoidal height $h$.
With semi-major axis $a$, eccentricity $e$, and prime-vertical radius
\[
N(\phi)=\frac{a}{\sqrt{1-e^2\sin^2\phi}},
\]
the Earth-centered Earth-fixed (ECEF) position is
\begin{align}
x&=(N+h)\cos\phi\cos\lambda,\\
y&=(N+h)\cos\phi\sin\lambda,\\
z&=(N(1-e^2)+h)\sin\phi.
\end{align}
Atlas retains these values in double precision. Regional tiles use an East-North-Up frame centered
at a geodetic origin. Render inputs subtract the camera in double precision before conversion to
float, avoiding centimeter-scale motion on top of million-meter coordinates.

The globe uses six cube faces. A normalized direction is projected to the dominant axis, producing
face-local $(u,v)\in[-1,1]^2$. A tile key stores face, level, integer $x,y$, and semantic layer.
Parent and four-child operations are exact integer transforms. Property and unit tests cover all
faces, geodetic round trips, local frames, poles, and invalid keys.

\section{Globe Selection}
Each frame begins with six root nodes. A tile's center direction maps to the WGS84 surface; corner
directions bound its angular radius. Atlas rejects a tile when its bounding sphere lies below the
ellipsoid horizon. Remaining spheres are tested against the camera view cone.

For geometric error $g$, distance $d$, viewport height $H$, and vertical field of view $\theta$,
screen-space error is
\[
\mathrm{SSE}=\frac{gH}{2d\tan(\theta/2)}.
\]
A priority queue refines the highest-utility node while respecting a hard tile-count bound.
Route-corridor nodes use a lower effective SSE threshold so limited refinement capacity reaches
future navigation content before unrelated detail. The selector records visited nodes, horizon and
view culls, refined nodes, route-biased nodes, and deepest selected level.

In a 640$\times$360 MoltenVK smoke run on an Apple M2 Pro, the bounded selector chose 382 leaves,
visited 562 nodes, horizon-culled 41 nodes, reached level 4, and marked a route-biased node. These
are execution diagnostics, not a global quality result.

\section{Streaming and Integrity}
The runtime accepts file, memory, HTTP, and SQLite regional-pack sources through one cancellable
asynchronous interface. HTTP supports ETags, conditional requests, timeout, redirect limits,
request cancellation, and byte-range content through the gateway. The disk cache performs atomic
temporary writes, quota-aware least-recently-used eviction, pinning, and metadata updates.

Every payload is hashed from its actual bytes using an internal SHA-256 implementation. Expected,
transport, cache, and pack checksums are compared against the calculated digest rather than merely
against each other. A corrupt cache entry becomes a miss; a corrupt pack row fails explicitly.
No unavailable tile or checksum error is silently accepted.

The `.vxa` format is a transactional SQLite archive. Its primary tile key contains layer, face,
level, $x$, and $y$; rows store payload, ETag, and SHA-256. Additional tables store metadata,
offline points of interest, and named routing/transit assets. Writers use a transaction, checkpoint
WAL state, and return to a portable delete journal before closing. Readers use independent
read-only connections.

The checked Connaught Place compatibility pack contains 453 GLB rows: 151 each at levels 18, 19,
and 20. The five reference manifests cover Delhi NCR, Greater London, Tokyo metro, New York metro,
and the Swiss Alps. They contain ten layer contracts and six implicit cube roots. CI regenerates
each manifest byte-for-byte.

\section{Navigation Contract}
Four C++ interfaces isolate consumer code from providers: \texttt{SearchProvider},
\texttt{RouteProvider}, \texttt{TransitProvider}, and \texttt{TrafficProvider}. The normalized
gateway exposes \texttt{/v1/search}, \texttt{/v1/reverse}, \texttt{/v1/route},
\texttt{/v1/transit}, \texttt{/v1/traffic}, \texttt{/v1/status}, and byte-range content.
\texttt{GatewayNavigationProvider} consumes this contract through libcurl.

A deterministic replay provider supplies place, road, transit, and traffic fixtures. The rendered
globe requests the replay route and constructs a tangent-plane ribbon above the ellipsoid.
Segments are geodesically subdivided before triangulation, so long routes follow the surface
instead of cutting through the globe.

The gateway never logs API keys. Its replay mode is a complete offline contract fixture. Self-hosted
deployments can place Pelias, Valhalla, OpenTripPlanner, traffic ingestion, and object storage
behind the same client-facing schema. Offline `.vxa` packs retain local POI search and can carry
regional graph extracts.

\section{Route-Predictive Controller}
For candidate tile $t$, route polyline $R$, and corridor radius $\sigma$, the controller first
computes the minimum ECEF distance $d(t,R)$ after subtracting the tile radius. Spatial route
probability is
\[
p_s(t)=\exp\left(-\frac{d(t,R)^2}{2\sigma^2}\right).
\]
Distance along the route and expected speed estimate seconds until needed $\tau(t)$. A bounded
look-ahead factor yields
\[
p_r(t)=p_s(t)\left(0.35+0.65\max\left(0,1-\frac{\tau(t)}{T}\right)\right).
\]

Visual utility combines semantic weight $w_s$, quality gain $Q$, visibility confidence $V$, SSE,
route probability, and deadline pressure:
\[
U(t)=w_sQ(0.25+V)+p_rW_{\rm route}W_{\rm deadline}
      +0.02\,\mathrm{SSE}.
\]
Cost includes one unit of fixed scheduling work, upload MiB, measured or estimated GPU geometry
cost, and lighting cost. Candidates are sorted deterministically by $U/C$ and tile key.

Selection enforces resident memory, per-frame upload, maximum changes, and speculative-waste
budgets. Frame-time pressure comes from an exponentially weighted moving average and reduces the
change limit above the target. Route deviation sets route probability to zero immediately.

Seven policies expose the controller's assumptions:
\begin{itemize}
\item distance-only: current visibility and geometric quality;
\item velocity-only: forward extrapolation from camera velocity;
\item route-only: route probability without semantic weights;
\item route-semantics: route and semantic weights;
\item geometry-only: lighting cost removed;
\item lighting-only: upload cost removed;
\item full Atlas: route, semantics, geometry, upload, and lighting.
\end{itemize}

\section{Renderer Integration}
Platform-specific Vulkan ownership is isolated behind \texttt{VulkanSurfaceHost}: framebuffer
extent, resize state, event waiting, instance extensions, and surface creation. Device and renderer
code no longer imports GLFW. The GLFW implementation remains the desktop host while Qt, Android
\texttt{ANativeWindow}, and headless hosts share the same contract.

Atlas is split into CMake libraries for core coordinates/data, network transport, navigation,
research control, rendering primitives, and streaming. \texttt{VulkaxAtlasCore} remains a
compatibility aggregate. The existing \texttt{LveEngine} executable preserves all BEACON and
GeoBEACON regression modes; \texttt{VulkaxAtlas} enables the globe entrypoint.

Offscreen exact-reference readback waits on the fence for the submitted frame. It no longer calls
\texttt{vkDeviceWaitIdle} inside the measured frame loop. Geo tile resources retire after more than
the maximum frames in flight; shutdown waits the graphics queue. Swapchain recreation and final
application shutdown retain explicit synchronization.

\section{Methodology}
The checked controller fixture contains 183 unique level-16 candidates around a 3.2 km Delhi route
and surrounding distractors. A 4 MiB resident budget holds only 64 tiles. Upload is limited to
4 MiB/s and two changes per frame; each tile is 64 KiB. Each policy receives the same 600-frame
fixed path, fixed $1/60$ s timestep, seed 1337, costs, route, and candidate set.

At each frame, the benchmark checks the current and three near-future route demands before applying
new decisions. A missing required key is a deadline miss. A resident tile never observed in the
route-demand set contributes to wasted prefetch bytes. The output includes every frame, cumulative
misses, selected and resident counts, upload and speculative bytes, semantic utility, route
quality, and budget violation.

This experiment isolates scheduling logic. Its measurement class is
\texttt{analytical-scheduler-simulation}; it must not be merged with Vulkan timings. Hardware
experiments use the separate renderer path, GPU query pools, deterministic cameras, warm-up
periods, exact diffuse captures, and raw CSV output.

\section{Results}
\begin{table}[h]
\centering
\small
\caption{Constrained 600-frame scheduler fixture.}
\begin{tabular}{lrr}
\toprule
Policy & Deadline misses & Wasted bytes\\
\midrule
Distance only & 644 & 3,145,728\\
Velocity only & 204 & 2,818,048\\
Route only & 8 & 2,686,976\\
Route + semantics & 4 & 2,686,976\\
Geometry only & 644 & 3,145,728\\
Lighting only & 644 & 3,145,728\\
Full Atlas & 4 & 2,686,976\\
\bottomrule
\end{tabular}
\end{table}

The fixture supports the preregistered existence claim: route-aware control beats distance-only
control on route-corridor misses under at least one constrained workload. Route-only accounts for
most of the reduction; semantic weighting resolves the remaining tie in this fixture. Velocity
prediction improves over distance but cannot anticipate turns as effectively as the recorded
route. Full Atlas and route-plus-semantics tie because this experiment does not execute actual
lighting or geometry kernels; that negative discrimination is expected and motivates hardware
coupled experiments.

The full policy also reduces wasted resident bytes by 458,752 (14.6\%) relative to distance-only.
The absolute values are fixture-dependent and are not presented as consumer-navigation performance.
Raw rows and the JSON summary regenerate the checked SVG figures without third-party plotting
packages.

\section{Verification}
The current automated suite covers BEACON bounds and scan behavior, WGS84/ECEF round trips, cube
addressing, manifest validation, SHA-256 vectors, cache and pack I/O, offline POI search, route
scheduling, runtime completion, replay and gateway providers, route mesh construction, bounded
globe selection, deterministic OSM tools, HTTP endpoint/range/ETag contracts, five reference
manifest regenerations, and the route-controller stress relation.

Linux CI builds release code, compiles shaders, runs CTest, builds the gateway container, resolves
Lavapipe explicitly, performs GeoBEACON and Atlas Vulkan smoke runs under Xvfb, and uploads raw
measurements. Local validation covers MoltenVK timestamp support and Vulkan CPU-fallback
classification on the second ICD.

\section{Limitations and Scope}
The checked five-region artifacts are manifests and deterministic source contracts rather than
committed full regional terrain databases; generated regional and planet payloads are intentionally
outside Git. Live data quality depends on configured OSM, terrain, GTFS, and traffic sources.
Traffic is optional BYOK and deterministic replay remains available.

The controller result reported here is analytical. Consumer usability, transit completeness,
cross-platform shell parity, and planet-scale service load require separate user, hardware, and
deployment studies. Formal BEACON error bounds exclude shadows, transparency, and strict specular
guarantees. Proprietary imagery, photogrammetry, and neural rendering are outside the artifact's
open procedural-data scope.

\section{Conclusion}
Vulkax Atlas turns the checked GeoBEACON renderer into a planet-coordinate, streamable, navigable
research substrate. Its central contribution is a transparent route-predictive controller that
jointly values future geometry, semantics, upload, memory, and lighting. The constrained fixture
demonstrates why route knowledge matters and identifies which components produce the effect.
Checksummed data boundaries, deterministic manifests, raw rows, explicit evidence classes, and
preserved negative cases keep the result reproducible and falsifiable.

\section{Artifact Reproduction Checklist}
The repository is designed to make each evidence class independently reproducible. A clean release
build first compiles the C++20 component libraries, the Vulkan application, \texttt{atlas-build},
\texttt{atlas-benchmark}, and every GLSL stage. CTest then executes core BEACON checks, Atlas
coordinate and scheduling checks, gateway contracts, manifest regeneration, GeoBEACON
preprocessing checks, and the constrained controller relation. The test runner fails if full Atlas
does not beat distance-only control in the preregistered constrained fixture.

Dataset reproduction begins from a region source manifest rather than from unrecorded local state.
For each of the five reference regions, \texttt{atlas-build generate-manifest} writes dataset
version 2 and a six-root implicit 3D Tiles descriptor. CI compares the generated bytes with the
checked files. Connaught Place pack reproduction starts from the committed OSM-derived GeoBEACON
manifest and GLBs, writes a transactional `.vxa` archive, checks SQLite integrity, and verifies 151
rows at each of three levels. Full generated regional and planet payloads remain external artifacts
because committing them would make normal source control unusable.

Research reproduction runs \texttt{atlas-benchmark} with frame count 600 and seed 1337. It writes
\texttt{frames.csv} before aggregation and records the evidence class in every row. The figure
script consumes only \texttt{summary.json} and emits the checked deadline-miss and wasted-prefetch
SVGs. No plotting library, network request, or GPU is required for this scheduler experiment.
Vulkan hardware measurements use the separate \texttt{VulkaxAtlas} or \texttt{LveEngine}
entrypoint and preserve device, driver, shader hash, timing source, resolution, warm-up count, and
camera path in the run manifest.

For visual verification, the Atlas entrypoint loads the deterministic road route, constructs the
ellipsoid-following ribbon, and displays permanent OSM attribution. The measured GeoBEACON path can
add exact and tested offscreen targets to the same submitted command buffer. Readback waits on that
frame's fence, writes reference/test PPM files, and reports actual image metrics separately from
the conservative diffuse bound. The final PDF is compiled from this checked \LaTeX{} source; page
count and rendered pages are inspected after compilation.

\clearpage
\begin{thebibliography}{9}
\bibitem{tiles} Open Geospatial Consortium, ``3D Tiles Specification 1.1,'' 2022.
\url{https://docs.ogc.org/cs/22-025r4/22-025r4.html}
\bibitem{osm} OpenStreetMap Foundation, ``Copyright and License.''
\url{https://www.openstreetmap.org/copyright}
\bibitem{replication} OpenStreetMap Wiki, ``Planet.osm diffs.''
\url{https://wiki.openstreetmap.org/wiki/Planet.osm/diffs}
\bibitem{pelias} Pelias, ``Open-source geocoder documentation.''
\url{https://github.com/pelias/documentation}
\bibitem{valhalla} Valhalla, ``Open-source routing engine.''
\url{https://valhalla.github.io/valhalla/}
\bibitem{otp} OpenTripPlanner, ``Open-source multimodal trip planner.''
\url{https://docs.opentripplanner.org/}
\bibitem{regir} Boksansky et al., ``Rendering Many Lights with Grid-Based Reservoirs,'' 2021.
\url{https://research.nvidia.com/labs/rtr/publication/boksansky2021rendering/}
\end{thebibliography}

\end{document}
